Verification of Stencil Computations through Spatial Specifications

### 2016-07-06 paper.tex
fix macro and some editing to intro

\newcommand{\trule}[1]{\footnotesize{#1}}

@@

\begin{abstract}
  Verifying the correctness of numerical computations relies on first
  giving a specification of their behaviour. In many cases, a
  numerical computation is derived from an underlying mathematical
  model, but this is often difficult to use as a specification as it
  is highly decoupled from the implementation. We have been studying
  specification techniques in middle ground between low-level and
  high-level specifications for numerical code. In this paper, we
  focus on \emph{stencil computations}---a common idiom in numerical code, but
  one which is error-prone due to fine-grained indexing errors. We
  hypothesise that in practise, stencil computations tend to have a
  regular shape which is amenable to static analysis and simple
  specification. We describe a abstract specification language for
  stencil computations that can be directly embedded as annotations in
  source code, along with an inference and checking procedure. We
  evaluate our language against a corpus of numerical Fortran code,
  and show that precise specifications can be given to roughly half of
  the identifiable stencil computations, and that the vast majority of
  stencil computations indeed have a simple, regular, static
  shape. This paper details our stencil specification language, its
  static semantics, inference, checking, and program synthesis
  procedures, our implementation, and our evaluation study. We provide
  an implementation of ontop of CamFort, an
  open-source program analysis and refactoring tool for Fortran corpus.
\end{abstract}

@@

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array value, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following iteration computes the one-dimensional discrete Laplace
transform (an approximation to a derivative) in Fortran:
%
\begin{minted}{fortran}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   end do
   a = b
end do
\end{minted}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 5 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. In Fortran, parentheses \texttt{( )} are used
for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.

%This stencil computation exhibits statically decideable
%spatial and temporal relationships between $\texttt{a}$ and
%$\texttt{b}$.
In this simple example, the dependency between \texttt{a}
and \texttt{b} on line 3 forms a simple spatial relationship which is easily
understood. This spatial relationship determines other aspects of the
program and its efficient implementation: how much ``boundary'' is
needed for the array, the most cache-efficient layout in memory,
the partitioning shape for parallel implementations.

More complex stencil computations can be much harder to understand and
subsequently more prone to error. For example,
Figure~\ref{ref:navier-stokes-fragment} shows three lines from a
Navier-Stokes fluid simulator in which two arrays are read from with
different data access patterns, across two dimensions. The interaction
is much harder to understand, with the potential for the developer to
accidentally introduce an error via simple textual mistakes, for
example writing $\texttt{(i-1,j)}$ instead of $\texttt{(i+1,j)}$.

@@

In this paper, we make the following contributions:
%
\begin{itemize}
\item we introduce a specification language for
stencil computations that captures many common forms
of data access pattern, both spatial and temporal (Section~\ref{sec:lang});
\item we detail inference and checking
algorithms for stencil specifications (Section~\ref{sec:analysis});
\item we evaluate our implementation of the approach
in the CamFort tool for Fortran verification, studying
a number of example programs to assess the usefulness
of this approach.
\dnote{insert results here}
\end{itemize}
%

>>>>

\newcommand{\trule}[1]{{\footnotesize{#1}}}

@@

\begin{abstract}
  Verifying the correctness of numerical computations relies on first
  giving a specification of their behaviour. In many cases, a
  numerical computation is derived from an underlying mathematical
  model, but this is often difficult to use as a specification as it
  is highly decoupled from the implementation. We have been studying
  specification techniques in the middle ground between low-level and
  high-level specifications for numerical code. In this paper, we
  focus on \emph{stencil computations}---a common idiom in numerical code, but
  one which is error-prone due to fine-grained indexing errors. We
  hypothesise that in practise, stencil computations tend to have a
  regular shape which is amenable to static analysis and simple
  specification. We describe a abstract spatial specification language for
  stencil computations that can be embedded as annotations in
  source code, along with an inference and checking procedure. We
  evaluate our language against a corpus of numerical Fortran code,
  and show that precise specifications can be given to roughly half of
  the identifiable stencil computations, and that the vast majority of
  stencil computations indeed have a simple, regular, static
  shape. This paper details our stencil specification language, its
  static semantics, inference, checking, and program synthesis
  procedures, our implementation, and our evaluation study. 
\end{abstract}

@@

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation outputs an array, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following computes the one-dimensional discrete Laplace
transform (an approximation to a derivative) in Fortran:
%
\begin{minted}{fortran}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   end do
   a = b
end do
\end{minted}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 5 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. In Fortran, parentheses \texttt{( )} are used
for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.

%This stencil computation exhibits statically decideable
%spatial and temporal relationships between $\texttt{a}$ and
%$\texttt{b}$.
The shape of the data access pattern on line 3 determines other aspects of the
program and its efficient implementation: how much ``boundary'' is
needed for the array, the most cache-efficient layout in memory,
and the partitioning shape for parallel implementations.
In this example, the access pattern is simple, static, and easily
understood. More complex stencil computations 
are much more prone to error from simple lexical mistakes and
harder to understand. For example,
Figure~\ref{ref:navier-stokes-fragment} shows three lines from a
Navier-Stokes fluid simulator in which two arrays are read from with
different data access patterns, across two dimensions. The interaction
is much harder to understand, with the potential for the developer to
accidentally introduce an error via simple textual mistakes, for
example writing $\texttt{(i-1,j)}$ instead of $\texttt{(i+1,j)}$.

@@

In this paper, we make the following contributions:
%
\begin{itemize}
\item we introduce a specification language for
stencil computations that captures many common forms
of data access pattern, both spatial and temporal (Section~\ref{sec:lang});
\item we detail inference and checking
algorithms for stencil specifications (Section~\ref{sec:analysis});
\item we evaluate our approach, providing an 
implementation as an extension of CamFort,
an open-source program analysis and refactoring tool for Fortran
corpus. We use this implementation to evaluate our approach against
a corpus of scientific applications, totalling roughly one million
lines of code, and we study some verification case studies.
\dnote{insert results here}
\end{itemize}
%
